Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-12-21 paint-contour-tree.tex
more changes for dcg

%\def\ProofInAppendix{TRUE}

\ifx\SoCGVer\undefined
   \newcommand{\InSoCGVer}[1]{}%
   \newcommand{\InNotSoCGVer}[1]{#1}%
\else
%   \def\ProofInAppendix{TRUE}
   \newcommand{\InSoCGVer}[1]{#1}%
   \newcommand{\InNotSoCGVer}[1]{}%
\fi

\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%

% \InNotSoCGVer{%
%    \newcommand{\SeeFullVer}{}%
%    \newcommand{\LemmaProofInFullVer}[1]%
%    {%
%       \begin{lemma}
%           #1
%       \end{lemma}%
%    }
% }
% 
% \InSoCGVer{%
%    \newcommand{\SeeFullVer}{{{\CiteFullVer}}}%
%    \newcommand{\LemmaProofInFullVer}[1]%
%    {%
%       \begin{lemma}[For proof see  \CiteFullVer{}]
%           #1
%       \end{lemma}
%    }   
% }

\InSoCGVer{%
   \documentclass[a4paper,UKenglish]{lipics-v2016}
%   \documentclass[11pt]{article}%

%  \usepackage{proofapnd}
}
\InNotSoCGVer{%
  \documentclass[11pt]{article}
  \usepackage[margin=2.54cm]{geometry}
}

\InNotSoCGVer{
   \newcommand{\myparagraph}[1]{\paragraph{#1}}
}
\InSoCGVer{
   \newcommand{\myparagraph}[1]{\smallskip\noindent{\textbf{#1}}}
}

@@

\InNotSoCGVer{
 \newtheorem{theorem}{Theorem}[section]
}
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\InNotSoCGVer{
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{example}[theorem]{Example}
 \newtheorem{definition}[theorem]{Definition}
}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
% \InNotSoCGVer{ 
%  \newtheorem{remark}[theorem]{Remark} 
% }
\newtheorem{Remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

@@

\InNotSoCGVer{
\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
  }
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}
}

\InSoCGVer{
% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\titlerunning{Avoiding the Global Sort: A Faster Contour Tree Algorithm} %optional, in case that the title is too long; the running title should fit into the top page column

\author[1]{Benjamin Raichel}
\author[2]{C. Seshadhri}
  \affil[1]{%
      Department of Computer Science, %
      University of Texas at Dallas\\ %
      Richardson, TX, 75080, USA\\ %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
   }%
   \affil[2]{%
      Department of Computer Science, %
      University of California, Santa Cruz\\ %
      Santa Cruz, CA, 95064, USA\\ %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
   }%

\authorrunning{B. Raichel and C. Seshadhri} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'

\Copyright{Benjamin Raichel and C. Seshadhri}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\subjclass{F.2.2, I.1.2, I.3.5}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
\keywords{contour trees, computational topology, computational geometry}% mandatory: Please provide 1-5 keywords
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\EventEditors{S\'andor Fekete and Anna Lubiw}
\EventNoEds{2}
\EventLongTitle{32nd International Symposium on Computational Geometry
(SoCG 2016)}
\EventShortTitle{SoCG 2016}
\EventAcronym{SoCG}
\EventYear{2016}
\EventDate{June 14-18, 2016}
\EventLocation{Boston, USA}
\EventLogo{}
\SeriesVolume{51}
\ArticleNo{57}

}

\maketitle
\InNotSoCGVer{
\thispagestyle{empty}
}

@@

\InNotSoCGVer{
\newpage
\pagenumbering{arabic}
}

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)

@@

\InSoCGVer{See the full version \CiteFullVer for further technical remarks on the above definitions.}
\newcommand{\technicalRemarks}{
%\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.
}
\InNotSoCGVer{
\subsection{Some remarks}
\technicalRemarks
}

@@

\newcommand{\newintuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a short, high-level description of the main ideas of subsequent sections. 
A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.
% However, for the $d=2$ case, 
% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
% can be explained clearly.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear.
% 
% 
% Each additional cut increases the complexity of the surface,
% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
% upwards and downwards.  We finally get a collection of extremum dominant pieces.
We prove a simple contour surgery theorem that builds the contour tree of $\MM$
from the contour trees of the various pieces created.

Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
which unlike the contour tree tracks superlevel sets rather than level sets.
% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
% except at a unique boundary face (think of a large boundary triangle that is the boundary).
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

@@

% What happens when finally blue and red join at $v$? We merge the two colors, but now have
% blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
% This necessitates using a priority queue with efficient merges.
% Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
% We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
}

@@

%Abusing notation, for now we will just call this the join tree. 
%(In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction between the join tree and the critical join tree.)
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

@@

\begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

% Relating the running time to a path decomposition is the most technical part of the paper. 
% All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
% for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
% to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
% Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
% saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
% higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
% before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.
% 
% Nonetheless, we can get a non-trivial (but non-optimal) bound, which is the analogue of \Thm{main-corr} for join trees. 
% Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
% The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
% $D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
% For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
% \Thm{main-alg}
%
% The cost of heap operations depends on heap sizes, which keep changing because
% of the repeated merging. This causes the analysis to be technically challenging. There are situations where
% the initial heap sizes are small, but they eventually merge to create larger heaps. 
% We employ a variant
% of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
% The final analysis charges expensive heap operations to long paths in the decomposition.

@@

Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
entropy argument, any algebraic decision tree that correctly computes the contour tree on all instances
requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
comparisons on all these instances. 
%In general, any algorithm requires $\Omega(C)$ time
%to solve the set of instances that our algorithm solves with $C$ comparisons.

@@

\InSoCGVer{

\newintuitionSection

% \subsection{Outline}
% The first step is to partition $\MM$ into nicely behaved pieces, such that the contour tree of
% each piece is a rooted tree.  This is done by cutting $\MM$ at certain contours. Of course, any such ``cut"
% increases the complexity of the input, so care is required to keep the complexity down.
% \Sec{surgery} describes our cutting procedure, and \Sec{rain} describes how to use this cutting procedure.
% It turns out that a simpler construct, the \emph{join tree}~\cite{csa-cctad-00} of each piece is exactly
% the contour tree. Thus, it suffices to 
% 
% Roughly speaking the resulting pieces are ``nice'' because the contour tree of each piece 
% looks like a rooted tree, i.e.\ the join tree of \cite{csa-cctad-00}, and the precise relation to join trees is the subject of \Sec{extreme}.
% \Sec{paint} then describes how to optimally compute the join tree (and hence the contour tree) 
% on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
% Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bounds the 
% running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
% in terms of path decompositions, and \Sec{lb} which provides matching running time lower bounds.
% 
% The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
% can be explained clearly. Therefore, in \Sec{approach} we provide a high level view of the entire result.
% We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
% as a simple matter of implementing these ideas.
% 
% Alternatively, the reader can view the full version of the paper \CiteFullVer{} which contains this high level overview and all other appendices 
% in the main body of the text.
% 
}

\InNotSoCGVer{
\intuitionSection
}

@@

\section{Divide and conquer through contour surgery} \label{sec:surgery}

\InSoCGVer{The following theorems are intuitively obvious, and are proven in the full version \CiteFullVer.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\surgeryBody}{
%To prove the above theorem, we require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
To prove Theorem~\ref{thm:surgery}, we require a theorem from \cite{c-tmi-04} (Theorems 6.6) that map paths in $\cC(\MM)$ to $\MM$.

>>>>

%\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%

  \documentclass[11pt]{article}
  \usepackage[margin=2.54cm]{geometry}

   \newcommand{\myparagraph}[1]{\paragraph{#1}}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}[theorem]{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
 \newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{example}[theorem]{Example}
 \newtheorem{definition}[theorem]{Definition}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
\newtheorem{Remark}[theorem]{Remark}
\newtheorem{observation}[theorem]{Observation}

@@

\author{
  Benjamin Raichel\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, 75080, USA; %
      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
  }
  \and
  C. Seshadhri\thanks{%
        Department of Computer Science; %
      University of California, Santa Cruz; %
      Santa Cruz, CA, 95064, USA; %
      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
  }
}

\title{%A Mountaintop View Requires Minimal Sorting:
Avoiding the Global Sort: 
\break A Faster Contour Tree Algorithm%
\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
}
\date{}

\maketitle
\thispagestyle{empty}

@@

\newpage
\pagenumbering{arabic}

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \Sec{lb}.)

@@

\subsection{Some remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees, and is the smallest tree that contains all the topological changes
of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

@@

%\pagebreak
\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a high-level description of the main ideas of subsequent sections.

\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
with exactly the same contour tree, but different orderings of (heights of) the critical points.
Turning it around, we cannot deduce the full height ordering of critical points from the contour tree, and so 
sorting all critical points is computationally unnecessary.
In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
another of the splits. Again, it is not necessary to know the relative ordering between the mounds
on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
information is necessary: on the left, the little valleys are higher than the big central valley,
and this is reflected in the contour tree. Leaf paths in the contour tree have points
in sorted order, but incomparable points in the tree are unconstrained.
How do we sort exactly what is required, without knowing the contour tree in advance?

\myparagraph{Join and split trees:}
The standard contour tree algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}, builds
two different trees, called the join and split trees, and then merges them together to get the contour tree.
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree, $\cJ(\MM)$, tracks exactly these events. 
Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
vertices higher than $v$. 
Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear (see \Sec{rain} for details).
% 
% 
% Each additional cut increases the complexity of the surface,
% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
% upwards and downwards.  We finally get a collection of extremum dominant pieces.
% We prove a simple contour surgery theorem that builds the contour tree of $\MM$
% from the contour trees of the various pieces created.

We now have carved $\MM$ up into a collection of extremum dominant pieces. 
In \Sec{surgery} we prove a simple contour surgery theorem allowing us to build the contour tree of $\MM$
from the contour trees of these various pieces.  Specifically, we argue our cutting in $\MM$ was done in such a way that 
each cut corresponds to cutting only a single edge in the contour tree.  Informally this means gluing $\MM$ back together along these 
cuts corresponds to gluing the contour tree together.

Ultimately, the reason for cutting $\MM$ into extremum dominant pieces is that such pieces have simpler contour tree structure. 
Specifically, using ideas from~\cite{csa-cctad-03}, in \Sec{extreme} we prove that the contour tree of an extremum dominant complex is (basically) just the join tree. 
Thus there is no longer a disconnect between the amount of sorting required for the contour tree versus the join tree, 
and so the problem of computing contour trees efficiently reduces to that for join trees.

\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
colors for each maximum. In arbitrary order, spill paint from each maximum, wait till it flows down,
then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
paint the interior of faces with dimension strictly larger than 1.} 
Furthermore, paints do not mix, so each edge receives a unique color, decided by the first paint to reach it.  
%
Note that the raining procedure, used above to carve the input into extremum dominant pieces, 
cannot be used here as it wets interiors of higher dimensional faces and so raining from each maximum may significantly increase the input complexity (see \Fig{order}).
On the other hand, the interface between two different colors is \emph{not} a contour, and so the join trees of each color class cannot just be simply glued together.

% In the following, $[n]$ denotes $\{1, \ldots, n\}$, for any natural number $n$.
% \begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
% A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
%  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
% 	consisting of edges in $E$, such that all edges along this path have the same color as $x$. 
% 
% An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
% \end{definition}

% Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
% of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
% colors are distinct. A few comments on paint vs water. The interface between two regions of different color
% is \emph{not} a contour, and so we cannot apply the divide-and-conquer
% approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
% Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
% Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
% dominant input, using water to divide the input $\NN$ raises the complexity too much.

Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum), by repeatedly merging colors and components together.  
We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
Let us focus on the initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
Both mounds are colored completely blue or red, while $v$ is touched by both colors.
So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this.
We know $u$ must also be touched by blue, so all critical vertices touched by blue
are put in a priority queue keyed by height, and vertices are handled in that order.

\myparagraph{Running time intuition:} 
 All non-heap operations can be easily bounded by $O(t\alpha(t) + N)$. 
 One can argue any heap always contains a subset of a single leaf to root path.  
 Bounding each heap by the size of this path suffices to prove the bound of \Thm{main-corr}, stated for join trees. 
 However, this may grossly overestimate heap sizes, as each vertex on a path can have at most two colors (the ones getting merged).
 Thus as we get closer to the root the competition for which two colors a vertex gets assigned to grows, since the number of descendant leaf colors grows. 
 This means for some vertices the size of the heap will be significantly smaller during an associated heap operation.

The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
%
This non-contiguous complication along with the fact that heap sizes keep changing as color classes merge, causes the analysis to be technically challenging. 
We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.

\newcommand{\newintuitionSection}{
\section{A tour of the new contour tree algorithm} \label{sec:approach}
We provide a short, high-level description of the main ideas of subsequent sections. 
A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.

\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
of the pieces is linear.
We prove a simple contour surgery theorem that builds the contour tree of $\MM$
from the contour trees of the various pieces created.

Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
which unlike the contour tree tracks superlevel sets rather than level sets.
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.25\linewidth]{figs/redBlue}%
    \hspace{.07in}
    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
    \hspace{.07in}
    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
    \hspace{.09in}
    \includegraphics[width=.3\linewidth]{figs/mound}%
    \hspace{.07in}
    \includegraphics[width=.065\linewidth]{figs/moundTree}
    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
    \label{fig:colors}
\end{figure}
}

@@

Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
are still put as vertices in the join tree. They simply form a long path. The split tree
is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.

@@

\begin{definition} \label{def:paint1} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$.

@@

@@

Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
entropy argument, any algebraic decision tree that correctly computes the contour tree on all instances
requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
comparisons on all these instances.

@@

@@

%\InSoCGVer{
%\newintuitionSection
%}

% \InNotSoCGVer{
% \intuitionSection
% }

\section{Painting to compute join trees}
\label{sec:paint}

We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
complex $\MM$.  In subsequent sections we then show how to reduce computing the contour to computing join by using contour surgery.

We now formally define the \emph{join} and \emph{split} trees as given by~\cite{csa-cctad-03}.
Conventionally, all edges are directed from higher to lower function value. 
In the following, for vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
numerous connected components.
